Burials and engravings in a small-brained hominin, Homo naledi, from the late Pleistocene: contexts and evolutionary implications

Version 1

Abstract

Abstract

Data from recent explorations in the Dinaledi subsystem illustrates one of the earliest examples of a mortuary practice in hominins and offers the earliest evidence of multiple interments and funerary actions, as well as evidence of the early creation of meaning making by a hominin. The hominin undertaking these behaviors was the small-brained Homo naledi. These data call into question several key assumptions about behavioral and cognitive evolution in Pleistocene hominins. The evidence from Dinaledi push back the temporal origins of mortuary and funerary behaviors and associate the creation of meaning making with a small-brained species and thus challenge key assumptions about the role and importance of encephalization in human evolution. This suggests that the hominin socio-cognitive niche and its relation to meaning-making activities is more diverse than previously thought. The association of these activities in subterranean spaces accessed and modified by the small brained species Homo naledi impacts assertations that technological and cognitive advances in human evolution are associated solely with the evolution of larger brains.

Main Text

Scholars have long argued that there is a qualitive difference between the abilities of Homo sapiens versus other Pleistocene hominins and that this difference has to do with overall brain size and neurobiological complexity. Complex behavior and the construction and use of “meaning-laden” material, e.g. burial, fire as light, engravings, ornamentation, ochre use, etc. have all been suggested to be signals of “modern” human cognitive capacity tied to a brain well above 1000cc (Kissel and Fuentes 2021; Galway-Witham, Cole, and Stringer 2019). The recent finds from the Dinaledi chamber, Rising Star Cave, South Africa indicate that large-brain-only model for complex hominin behavior no longer holds (Berger et al. 2023a,b).

It is true that overall brain size and Encephalization Quotient (EQ) increased in many populations of the genus Homo over the past 2 million years, and that this is generally associated with increased appearance of complex behavior. However, it is now apparent that many of these behaviors are found in places used by, or associated with, hominins that are not traditionally considered to be Homo sapiens. For example, engravings are found with non-H. sapiens hominins (Joordens et al. 2014; Mania and Mania 1988; Sirakov et al. 2010). Ochre use is documented in samples that predate H. sapiens (Ronen et al. 1998; Watts, Chazan, and Wilkins 2016; Dapschauskas et al. 2022) and both fire use and simple mortuary behavior are also found in association with hominins that pre-date H. sapiens (Carbonell and Mosquera 2006; MacDonald et al. 2021). However, one might argue that all of these instances are associated with large-brain hominins or could be the result of taphonomic issues or dating issues (Püschel et al. 2021). But this is not accurate. Substantive evidence suggests that approximately 250-350,000 years ago the later Pleistocene hominin, Homo naledi, a small-brained hominin, carried deceased conspecifics into difficult to access locations in the Rising Star Cave system and interred them (Berger et al. 2023 a,b) and likely produced engravings on cave walls near those areas of interment. Such actions require considerable social collaboration, coordination, and planning. And in the context of the subterranean Dinaledi system also require use of a light source; fire (Berger 2022). H. naledi carried out these behaviors with a brain size of less than ∼600cc.

Fire use, mortuary behavior, and the evidence of engravings attributed to H. naledi falsify the hypothesis that only a large-brained hominin was capable of cognitively complex cultural, possibly symbolic, behavior. Recent discoveries demonstrate that at least a few populations/taxa in the genus were characterized by smaller overall brain sizes (∼5-600cc) well into the later Pleistocene. Material evidence associated with these smaller brained populations overlaps with the technology and fire use (H. floresiensis, (Moore et al. 2009)), mortuary behavior and the production of engravings (H. naledi, Berger et al. 2023a,b) exhibited by their larger brained congenerics (see Fig 1). While these populations were characterized by smaller bodies than temporally sympatric hominins, their brains are not simply allometrically-scaled reductions of the larger forms (Holloway et al. 2018). This suggests that neurobiological organization rather than overall brain size, may have been one part of an early key transition within hominin evolution (8). It may be possible that the apparent gap in distribution of Homo/hominin fossils with an endocranial volume under 800cc between 1.0 and .3 mya in Figure 2 may be filled in with future discoveries. The open question introduced by the current evidence for H. naledi, and H. floresiensis, behavior is whether there might be one or more lineages of small brained hominins that flourished alongside larger brained forms and shared at least some aspects of the distinctive socio-cognitive hominin niche of the later Pleistocene (Galway-Witham, Cole, and Stringer 2019; Kissel and Fuentes 2021).

Archaeological evidence of culturally-mediated, meaning-making, behaviors.

Dots represent different sites and the error bars are the maximum and minimum dates when available. See Table for details and See supplemental material for references.

Endocranial volume estimates for hominin cranium.

Error bars represent the maximum and minimum ages for specimens when available. See supplemental material for references. Hawks, John. 2023. Endocranial volumes for fossil hominins (dataset). Figshare https://doi.org/10.6084/m9.figshare.22743980

Here we offer an analysis of the recent data reported for complex behavior in the small-brained Homo naledi and suggest a suite of implications this has for our understanding of the relationships between brain size, cognition, complex behavior and the evolution of the genus Homo across the Pleistocene. These implications also query the driving forces behind encephalization and its relationship to the emergence of complex behaviors in hominins and other animals.

H. naledi buried their dead

Recently Pettitt (2022) laid out the three key criteria for assessing whether or not the Homo naledi remains in the Dinaledi subsystem represent actual funerary behavior: a) is there an as-yet unmapped entrance into the Dinaledi Chamber? b) Is there any evidence of artificial lighting in the cave system, and c) Is there evidence that it was dead bodies, rather than body parts that were carried into the chamber? The first query has been repeatedly addressed and no other options for alternative ingress have been found or hypothesized (Elliott et al. 2021). And now Berger et al. (2023a,b) answers both (b) and (c) clearly: there is evidence of fire use in the Dinaledi system (Bower, 2022) and entire bodies rather than body parts make up much of the buried remains. Therefore, one must conclude that the evidence meets Pettit’s (2022) criterion set forth and that H. naledi remains in the Dinaledi subsystem are one of the two earliest examples of a mortuary practice and offer the earliest evidence of multiple interments and funerary actions in a hominin.

The locations, contexts, and the inferred behavior associated with the H. naledi burials, including the appearance of engravings near the interment sites, likely also demonstrate meaning-making activity (Kissel and Fuentes 2017). Mortuary and funerary practices had previously only been attributed to Homo sapiens but more recently are associated with a range of large-brained hominin taxa in later the Pleistocene (Figure 1). Evidence of funerary behavior is assumed to require human-like cognitive capability (Pettitt 2018). Such behavior found in a small-brained hominin suggests that increases in brain size/EQ may not be a necessary precursor for the appearance of meaning-making behavior in the hominins.

The newly described Homo naledi data, and the clear assignation of funerary behavior to the context, and the presence of engravings in the same space, calls into question several key assumptions about behavioral evolution in Pleistocene hominins, and the importance of brain size evolution in general. The data emerging from the Dinaledi subsystem support an emerging argument that individual cognitive ability related to increased encephalization may not be the primary explanation of certain complex hominin behavior, such as burying of the dead, strategic use of fire for illumination, and the creation of engravings. Rather than wholly relying on increased encephalization, we suggest that a distinctive cultural, empathetic, collaborative niche dependent on increasingly complex and robust relationships between individuals has also been a primary driver in the development of key aspects of human, or human-like, behavior (Galway-Witham, Cole, and Stringer 2019; Kissel and Fuentes 2021; McBrearty and Brooks 2000; Fuentes 2017; Spikins 2022; DeCasien, Barton, and Higham 2022)

The varied treatments of the dead described in Berger et al. (2023a) is certainly indicative of mortuary behavior, defined as specific actions relating to death and to the treatment of the dead, and the presence of dug burials and associated engravings is likely indicative of funerary behavior, defined as specific activities relating to the disposal of the dead and to their subsequent commemoration (Pettitt 2018). The data reflect a scenario where members of the H. naledi community carried the bodies of their dead more than 75 meters underground in an extremely difficult and dangerous subterranean environment (Elliott et al. 2021). Despite bodies being smaller and thus more capable of navigating the Rising Star system than contemporary humans, the carrying of bodies and behavioral actions required to bury or place them in the various locations described would have had high energetic costs and carried substantial risk for H. naledi. There are no clear direct fitness benefits nor any indication of proximate functional stimuli for this suite of behaviors.

The challenges for Homo naledi to bury its dead in remote subterranean contexts are significant. To accomplish this H. naledi had to coordinate their behavior and collaborate to move the bodies to a specific location inside the cave system, excavate an area and place the body in it, and cover it, or place the remains in a specific non-floor area. The data from Berger et al. (2023a) suggest that this is a behavioral sequence that was repeated multiple times in the same location, likely across a long temporal duration. Given the complexity of the cave layout (Elliott et al. 2021), there must have been some form of explicit communication for coordination of movement and actions, and the use of fire as a light source, between the H. naledi undertaking the behavior. Such coordination and specific set of actions around the treatment of deceased conspecifics is more methodologically extensive, energetically costly, with higher risk of injury than any reported for other primates and non-human animals to date (King 2013). This behavior is also more complex and multifactorial than that reported for the one earlier case of hominin mortuary behavior (Sima de Los Huesos, (Carbonell and Mosquera 2006)).

The subterranean environment used by H. naledi is not only physically challenging but is also emotionally and physiologically challenging, reflecting an engagement with difficult underground spaces not seen elsewhere in the archaeological record at this time. Dark enclosed spaces, where visual perception is curtailed, can create a state of emotional arousal profoundly affecting perceptual, cognitive and social systems (Zuccarelli et al. 2019), even with some form of illumination. While we cannot yet be certain of the exact modes, intensity and quality of the fires used by H. naledi in the Dinaledi subsystem, we can assume that they likely provided flickering and, at best, moderate intensity light sources. In humans, and other diurnal primates, sensory deprivation through reduced or a lack of consistent visual clues creates a heightened sensitivity to other senses as well as prompting experiences of visual disturbances, hallucinations, and disorientation (Hodgson 2021). Experiences of these types of extreme and unusual environments, though often inducing fear responses, can also facilitate powerful bonding experiences (Steidle, Hanke, and Werth 2013). This range of substantive emotional and psychological reactions may explain why experiences in deep dark caves are often associated with a sense of the transcendent in contemporary humans (Montello and Moyes 2012) and given the broad range of sensory commonalities across diurnal anthropoids, and especially apes, they likely had comparable impacts on H. naledi and other Pleistocene hominins using subterranean spaces. We argue that careful and coordinated treatment of the dead on several occasions, in these environments, implies particularly strong social and emotional bonds and some shared understanding of meaning (Pettitt and Anderson 2020) in H. naledi.

That this high-risk, high-cost, no-overt-direct-fitness-benefit behavior was undertaken repeatedly by multiple members of the H. naledi community across time indicates a valued social, and likely cultural, tradition. The combination of features of the behavior (burial and associated engravings), and the context in which it was undertaken (in deep caves with the use of fire for illumination), suggests a level of cognitive/semiotic meaning-making capacity in H. naledi (e.g., (Kissel and Fuentes 2017; 2018)) that matches similar assessments of other populations of the genus Homo during the same, and later, time periods (Figure 1). This assessment of meaning-making capacity for H. naledi is supported by the active transformation of the ‘space’ of the Dinaledi chamber and Hill antechamber to ‘place’ (Low and Lawrence-Zúñiga 2003) through the pattern of mortuary and likely funerary behavior (e.g. (Silverman 2008) involving group collective practice, high levels of social/emotion investment, and some form of communicating the commitment to undertake the endeavor.

Recent research has produced growing evidence of multiple aspects of meaning making earlier in the archaeological record and with a wider range of hominins acknowledged to be responsible for mortuary practices (Table 1 & Fig. 1). This includes meaning making within deep caves by members of the genus Homo beginning after 200,000 years ago at Wonderwerk in South Africa and Bruniquel in France, both of which have evidence of fire use as well. The range of examples of meaning making extend our understanding of the cultural complexity of later Pleistocene members of the genus Homo. Nonetheless, the Dinaledi subsystem site is the earliest known example of the pattern of mortuary, and likely funerary, behavior that becomes increasingly common in populations of the genus Homo in the terminal Pleistocene (Table 1). It is critical to note that H. naledi is the least Homo sapiens-like of any hominin yet described in the Middle and Late Pleistocene of Africa.

A role for emotional cognition?

This pattern of emergence of meaning making in different hominins (Figure 1), to which the Homo naledi burial evidence makes an important contribution, also has other implications. Most particularly it may cast light on the question of the evolution of human conscious awareness of emotions and its involvement in the processes of cognitive evolution in the genus Homo and in hominins in general. Humans share many building blocks of emotional cognition with other mammals, and some complex abilities with other apes. We share the same visceromotor and sensorimotor foundation for emotions with other mammals for example (Steklis and Lane 2013). Moreover, a range of common emotional responses to humans has now been documented in other primates, and particularly in apes, through measurement of heart rate and skin conductance as well as more recently pupil mimicry and infrared thermography (Nieuwburg, Ploeger, and Kret 2021). Interpersonal emotional interactions also have a common basis. Emotional contagion is apparent in monkeys and apes, and apes in particular demonstrating a level of empathy though yawning and even sympathy through active consolation (Romero, Castellanos, and de Waal 2010; Preston and de Waal 2002). Diverse primate species have the cognitive ability to infer emotional meaning from expressions (Nieuwburg, Ploeger, and Kret 2021). Moreover, there is anecdotal evidence for the foundations of cognitive empathy in targeted helping within apes (Koski and Sterck 2010). However, no other mammal demonstrates the emotional conscious awareness of contemporary Homo sapiens. Contemporary humans possess a distinctive cognitive ability to regulate emotions by bringing feelings into “rational” thought (Green and Spikins 2020). As a result humans communicate and engage in shared intentions, and meaning-making, to a degree not seen in other animals, and demonstrate motivations to share emotions, experiences and activities with other persons (Steklis and Lane 2013).

The H. naledi evidence suggests that conscious emotional awareness was present in this hominin despite its small brain size. The shared and planned deposition of several bodies in the Rising Star system is evidence of a shared sets of beliefs or assumptions about meaning and action, something one would term “shared grief” in contemporary humans. The creation of engravings in the same space as the burials suggests a form of shared memorialization, or at least shared attentions and action to alter the locations beyond the interment of the bodies. Regardless of what one terms the underlying cognitive processes associated with the burial activities of H. naledi, they indicate a level of conscious emotional awareness that enables and is associated with extensive shared intentionality, forward planning, and repeated cultural behavior involving bodily risk. Equally complex symbolic use of caves by Neanderthals (Jaubert et al. 2016; Baquedano et al. 2023) demonstrate a similar emotional self-awareness, and production of highly symmetrical stone tools is also potentially indicative of emotional awareness and regulation in early member of the genus Homo (Green and Spikins 2020). However, the fact that a small-brained hominin displays these sorts of behavior suggests that the neurological capacity enabled by a larger than 1000cc brain cannot be the only factor, or necessarily the main factor, enabling the kind of emotional cognition that is considered a central factor in human evolutionary success.

That complex emotional cognition is not unique to Homo sapiens should not be surprising. Social understanding of emotions is widely accepted as adaptive in an evolutionary context (Nieuwburg, Ploeger, and Kret 2021) and moreover emotional awareness is associated with better life outcomes in modern human contexts (Smith et al. 2023). That brain regions associated with socio-emotional processing are relatively enlarged in H. naledi despite a limited overall brain size reinforces this argument, particularly as the medial prefrontal cortex is important in reflective awareness in modern humans. A specifically H. naledi behavioral adaptation, reflected in the burial activities, may have depended on emotional commitments to others combined with a set of cultural beliefs/practices, a high level of emotional awareness to manage these, and in turn collaboration with extensive coordination.

What does H. naledi burial activity mean for understanding human evolution?

As outlined in Hawks et al. (Hawks et al. 2017) the phylogenetic relationships between H. naledi and other populations/taxa in the genus Homo remain unclear. However, H. naledi and some other populations of the genus Homo overlap temporally in the expression of meaning making behavior indicating some degree of shared socio/emotional/cognitive processes. To date, H. naledi is the earliest example of such actions, combining both mortuary and funerary behavior with the creation of likely symbolic engravings. Thus, it is clear that the hominins in the later Pleistocene are typified by a range of brain sizes and cranial and post-cranial morphologies and that the material record in that same time period offers increased evidence for shared meaning-making. This demonstrates that such behavior is neither “modern” nor exclusive to Homo sapiens (sensu latu). Whilst this adds further evidence to our understanding of the emergence of hominin cognition there are also wider evolutionary implications. Much like potentially independent evolution of social emotional abilities in other primates (Nieuwburg, Ploeger, and Kret 2021) the behavioral evidence for small-brained H. naledi may suggest that some degree of analogous rather than homologous evolution underlies social emotional complexity in humans.

The evidence for the burials by H. naledi creates two problems for current models of human evolution. The first is: what is the relationships between H. naledi mortuary/funerary behavior and that of other Homo taxa/populations? This can be clarified via one of three possible explanations.

H. naledi is not in the human lineage and its mortuary/funerary activities and use of fire are the result of parallel evolution/homoplasy.

mortuary practices, use of fire and any related meaning-making capacities are very old dating to the early Pleistocene, or even Pliocene hominins, and thus are a homology between H. naledi and H. sapiens. or,

H. sapiens “borrowed” the mortuary and funerary behavior as a cultural practice from sympatric H. naledi or vice versa.

The second problem is how to incorporate the fact that neither absolute brain size nor encephalization quotient are necessarily correlated with the meaning-making capacities and emotional-cognition complexity associated with mortuary and funerary behavior. It is assumed that a large brain was an essential step towards a uniquely human cognition, social relationships and culture (Dunbar 2003; Muthukrishna et al. 2018). However, small-brained hominins were responsible for many key changes in human evolution. Planning and forethought in stone tool production predates the origins of Homo (Harmand et al. 2015) and by 1.76 million years ago multiple taxa/populations of relatively small-brained hominins were likely developing separate bifacial tool traditions (Lepre et al. 2011). It is also evident that small-brained hominins (under 800-1000cc) were those who initially expanded around, and out of Africa, crossing into eastern and Southeastern Asia (Antón, Potts, and Aiello 2014). Additionally, the use of fire emerges in excess of 1.5 million years ago conspecific only with small-brained hominins (Hlubik et al. 2019). This constellation of data is particularly significant as it demonstrates that small-brained hominins are a part of the complex hominin niche that characterizes later Pleistocene members of the genus Homo (Mondanaro et al. 2020).

This new evidence for complexity in H. naledi behavior pushes back the origins of mortuary and funerary behaviors, challenges our assumptions about the role and importance of encephalization in human evolution, and suggests that the hominin emotional, socio-cognitive niche is more significant than previously thought. This will structure how we understand and model the origins and patterns of human evolution in the future (Kissel and Fuentes 2021; Spikins 2022; Spikins et al. 2019)

eLife Assessment

The discovery of Homo naledi-associated evidence for intentional burial and engravings would undoubtedly have important implications for our understanding of the evolution of complex cognition and behavior. Based on claims made in two related preprints by Berger et al., this study discusses the potential implications of the purported mortuary practice and symbolic behaviors claimed to be associated with the small-brained H. naledi remains. Unfortunately, the evidence presented in the two related submissions that the current paper entirely relies on is incomplete at this stage.

Reviewer #1 (Public Review):

As this experience as a reviewer has been unusual, it may be helpful to outline some relevant parameters of the task at the outset. While I was invited to review the Fuentes et al. study only, two additional papers concerning the claimed engravings and burials associated with Homo naledi by Berger and colleagues were also provided as components of the reviewer package. The two manuscripts presenting the archaeological evidence are accessible as preprints in bioRxiv, by Lee Berger and colleagues ('2023a, 2023b').

Unfortunately, the arguments in the Fuentes et al manuscript hinge entirely on the strength of archaeological evidence for engravings and intentional burial by Homo naledi (presented in the abovementioned two preprints). All inferences regarding hominin behaviour and biology of Homo naledi, discussed by Fuentes and colleagues, are wholly dependent on the evidence presented in the archaeology preprints being true.

Yet both of the archaeological manuscripts are unfortunately weak. In short, the claims for engravings depend on the demonstration of several elements of association that are rather standard for linking material traces found in the archaeological record with particular hominin behaviours. For the particular arguments by Berger and colleagues to be demonstrated, the traces on the rock surface need to be linked causally with hominin agency, in other words, their anthropogenic nature need to be established. The author of the engravings needs to be demonstrated as a particular hominin species (Homo naledi in this case), and the activity of engraving needs to have taken place ~241-335 kya. After reading the manuscript on the engravings, however, what is clear is that the scratches could as easily have been made by a modern-day farmer 50 years ago, as Homo naledi ~335 kya. Berger and colleagues do not present any evidence to the contrary, they simply describe their narrative as the most parsimonious scenario. A particularly curious piece of information presented as evidence is a list of individuals known to have entered the Dinaledi system in recent times (and known not to have scratched the walls, one presumes, though this is not stated).

The question of intentional burial is more complex. What we know from other widely accepted early burials is that documenting the geoarchaeological context of the hominin remains is critical to assess the likelihood of an intentional burial - this needs to be established at the outset through high quality fieldwork. Yet even the boundaries of the excavation presented in the burial manuscript appear so angled or skew relative to one another (Fig. 2a) that the individual squares look to be aligned with different XY grids, which does not instill confidence in the quality of field documentation. One can make out very little from the sediment section images - which are key to identifying intrusive features associated with burials - and the multivariate geochemical analysis of sediments is unconvincing: a scatterplot (not a biplot) should have been provided showing the geochemistry of the burial sediment samples relative to the immediately surrounding sediment characteristics. While one remains excited about the potential for a spectacular archaeological discovery within the Dinaledi cave system, unfortunately, the three manuscripts provided do not present convincing evidence to that effect.

Reviewer #2 (Public Review):

Fuentes et al. provide a detailed and thoughtful commentary on the evolutionary and behavioral implications of complex behaviors associated with a small-brained hominin, Homo naledi. Within the Rising Star Cave of South Africa, Berger et al. 2023a,b proposed evidence that Homo naledi intentionally buried their dead through complex mortuary practices and engaged in symbolic expression by engraving the cave walls in cross-hatching motifs. Two burials were identified in the Rising Star cave subsystems: Feature 1 in the Dinaledi Chamber and a feature in the Hill Antechamber. The engravings are located in the Hill Antechamber near the passageway leading into the Dinaledi chamber. The authors aimed to provide evidence for burials by (1) testing sediment samples for mineral composition from within and outside the burial feature; (2) demonstrating an interruption in the stratigraphy indicative of a "bowl-shaped" feature; (3) evaluating the anatomical coherence of the skeletal remains; (4) demonstrate matrix-supported positioning of skeletal elements; and (5) determine the compatibility of non-articulated material with decomposition and subsequent collapse. Berger et al. 2023b evaluated the engravings through high resolution photography, cross-polarization, and 3D photogrammetry. Neither article involved radiometric dating of materials. While the review by Fuentes et al. highlights important assumptions about the relationship between hominin brain size, cognition, and complex behaviors, the evidence presented by Berger et al. 2023a,b does not support the claim that Homo naledi engaged in burial practices or symbolic expression through wall engravings.

The major weaknesses for Berger et al. 2023a are as follows:

1. The mineral composition from sediment sampled from within Dinaledi Feature 1 is not different compared to the surrounding sediment, which is one rationale proposed by the authors that would lead to the conclusion of a burial pit. An effort to replicate the multivariate statistical analysis using the data provided in SI Table 1 by this reviewer failed, and thus, the results are not replicable.

2. The authors failed to provide clear visualizations or analysis that showed an unambiguous interruption in the stratigraphy surrounding the Dinaledi Feature 1.

3. Attempts 1 and 2 were applied solely to Dinaledi Feature 1, not the Hill Antechamber Feature.

4. Skeletal cohesion does suggest that the bodies were likely covered or protected by external environment. However, given the geological context, there is minimal opportunity for scavengers or other agents to scatter the skeletal remains within such an isolated location. Thus, this alone cannot solely support intentional burials as this line of evidence is subject to equifinality.

5. Similar to the preceding statement, evidence for matrix-supported elements was inconclusive at best. There was no mention of sedimentary rate or expectations for how quickly sediments would naturally bury the remains of whole bodies in the chamber compared with the rate of decomposition of buried remains.

The major weaknesses for Berger et al. 2023b are as follows:

1. While this is incredibly difficult to accomplish, dating rock art or other cave wall engravings is the only method to ensure that the etchings were created during the time of Homo naledi. Unfortunately, this was not attempted. Instead, the authors state that "This description is intended to document the discovery and provide spatial and contextual information prior to any further analyses that may require invasive sampling." Yet, the authors assign a date to the engravings in the title of the paper. Here, the authors are generating interpretations before analyses are attempted.

2. The engravings are indeed very interesting and are likely anthropogenic in origin. However, the argument that these engravings were created by Homo naledi is based on the bold assumption that "No physical or cultural evidence of any other hominin population occurs within this part of the cave system, and there is no evidence that recent humans or earlier hominins ever entered any adjacent area of the cave until surveys by human cave explorers during the last 40 years." (page 6). To assume that no other individual entered the cave system from the time of Homo naledi until 40 years ago is an unrealistic and faulty assumption. This reviewer does not discount that the engravings could have been made by Homo naledi, but the evidence must be sufficient to support this statement or provide other alternatives as working hypotheses.

As a discipline, paleoanthropology aims to understand the evolutionary history of the hominin clade through fossil remains, material culture, and, most recently, ancient DNA. The methods and approaches that we as paleoanthropologists use to understand the past often bridge both the humanities and the hard sciences to create a unique understanding of our shared history. We are only limited by the conditions in which time and attrition has erased pieces of our collective story from the earth. Thus, it is our responsibility to ensure that our interpretations of the past are supported by measurable and testable means, to the best of our ability, and that hypotheses are not presented as conclusions.

Unfortunately, this is not the case for Berger et al. 2023a,b. The work presented by the authors is imprudent and incomplete and does not meet the requirements set forth by our discipline. While it is important that scholars publish their work in a dutiful timeline, it is arguably more critical for scholars to take the necessary time to ensure the integrity and resolution of the work. The consequences for rushing publications with such a significant unsubstantiated find will likely result in perilous ramifications, as it is more difficult to correct an idea than to introduce one.

Reviewer #3 (Public Review):

This paper presents the cognitive implications of claims made in two accompanying papers (Berger et al. 2023a, 2023b) about the creation of rock engravings, the intentional disposal of the dead, and fire use by Homo naledi. The importance of the paper, therefore, relies on the validity of the claims for the presence of socio-culturally complex and cognitively demanding behaviors that are presented in the associated papers. Given the archaeological, hominin, and taphonomic analyses in the associated papers are not adequate to enable the exceptional claims for naledi-associated complex behaviors, the inferences made in this paper are currently inadequate and incomplete.

The claimed behaviors are widely recognized as complex and even quintessential to Homo sapiens. The implications of their unequivocal association with such a small-brained Middle Pleistocene hominin are thus far reaching. Accordingly, the main thrust of the paper is to highlight that greater cognition and complex socio-cultural behaviors were not necessarily associated with a positively encephalized brain. This argument begs the obvious question of whether absolute brain size and/or encephalization quotient (i.e., the actual brain volume of a given species relative the expected brain size for a species of the same average body size) can measure cognitive capacity and the complexity of socio-cultural behaviors among late Middle Pleistocene hominins.

Claims for a positive correlation between absolute and/or relative brain size and cognitive ability are not common in discussions surrounding the evolution of Middle- and Late Pleistocene hominin behavior. Currently, the bulk of the evidence for early complex technological and social behaviors derives from multiple sites across South Africa and postdates the emergence of H. sapiens by more than 100,000 years. Such lag in the expression of complex technologies and behaviors within our species renders the brain size-implies-cognitive capacity argument moot. Instead, a rich body of research over the past several decades has focused on aspects related to socio-cultural, environmental, and even the wiring of the brain in order to understand factors underlying the expression of the capacity for greater behavioral variability. In this regard, even if the claimed evidence for complex behaviors among the small-brained naledi populations proves valid, the exploration of the specific/potential socio-cultural, neuro-structural, ecological and other factors will be more informative than the emphasis on absolute/relative brain size.

The paper presents as supporting evidence previous claims for the appearance of similar complex behaviors predating the emergence of our species, H. sapiens, although it does acknowledge their controversial nature. It then uses the current claims for the association of such behaviors with H. naledi as decisive. Given the inadequate analyses in the accompanying papers and the lack of evidence for stone tools in the naledi sites, the present claims for the expression of culturally and symbolically mediated behaviors by this small-brained hominin must be adequately established. The importance of the paper thus rests on the validity of the claimed evidence--including contextual aspects--for rock engraving, mortuary practices, and the use of fire presented in the associated two papers. The claims in both associated papers are inadequate, incomplete, and largely assumption- (rather than evidence) based. As responsible and ethical researchers, the team must return to the sites, conduct the required standard chronomoetric and taphonomic studies and weigh the strength of the evidence before proceeding with the current claims.

Author response:

The following is the authors’ response to the current reviews.

We thank the editors at eLife and the one reviewer for engaging our revised manuscript. As we noted in our previous response to reviewers, which we wrote in October 2024 when we submitted our initial revision the majority of critique we received was targeted not so much at the argument of this manuscript but at the debate regarding the evidence in the two other manuscripts that this one accompanied; “ Evidence for deliberate burial of the dead by Homo naledi” and “241,000 to 335,000 Years Old Rock Engravings Made by Homo naledi in the Rising Star Cave system, South Africa.” Because of that critique we revised this manuscript to emphasize that the key element in constructing our argument is that *H. naledi *engaged in mortuary behavior (the movement of dead H. naledi by living H. naledi into the Rising Star cave system) and place that in context of a) the increasingly complex later Pleistocene record of meaning making activity and b) the assumed correlations between brain size and cognitive capacities in Pliocene and Pleistocene hominins. This framing, as noted in the eLife editorial comment, is the main thrust of our manuscript. There is a growing convergence of evidence that totality of the currently available data and analyses for H. naledi in the Rising Star cave system support mortuary behavior: that is, the agential and intentional action by H. naledi individuals in the transport of bodies to the Lesedi Chamber and Dinaledi Subsystem--see Berger et al. 2025 plus the 2nd round reviews and the eLife editorial comment associated with it, and also Van Rooyen et al. 2025. We acknowledge the serious debates around the assertion of funerary behavior (cultural burial) and seek to illustrate that while we believe the data support the funerary behavior hypothesis, it is not a necessary requirement for our main argument.

A few specific responses to the reviewer in this revised manuscript:

Reviewer states: “Claims for a positive correlation between absolute and/or relative brain size and cognitive ability are not common in discussions surrounding the evolution of Middle- and Late Pleistocene hominin behavior.” We are not making the argument that absolute brain size in the later Pleistocene is a point of focus, rather that there are many arguments and assertions about EQ and cognitive capacity that are central in the proposals for the evolution of hominins in general and genus Homo in particular across the Plio-pleistocene period. We offer a brief review of this in the text and suggest, as noted by this reviewer, that “exploration of the specific/potential socio-cultural, neuro-structural, ecological and other factors will be more informative than the emphasis on absolute/relative brain size”…this (in their words) is exactly our main point. However, we contend that such a framing should not be exclusive to later Pleistocene contexts, but rather that the examination of earlier hominins might also be better served by moving away from the traditional assumptions of cognitive complexity associated with absolute/relative brain size. The reviewer states: “The authors use, in a number of instances throughout the paper, secondary sources of information such as review papers (e.g., McBrearty & Brooks 2000; Scerri & Will 2023; Galway-Witham et al. 2019) instead of the original works that are the basis for making the desired case.” We do indeed use review papers in the main body of the text for clarity, brevity, and to acknowledge robust previous review work in these areas, however in the supplemental text and with the figures and table we offer substantive bibliographies of the original citations and studies. We encourage readers to please spend time with those materials as well. Finally, the reviewer states: “Given the inadequate analyses in the accompanying papers, and the lack of evidence for stone tools in the naledi sites, the present claims for the expression of culturally and symbolically mediated behaviors by this small-brained hominin must be adequately established.” We are quite specific in this manuscript, and in other publications, that we are not arguing for “symbolically mediated” behavior, but do stand by our non-controversial suggestions of meaning-making, and cultural behavior, as relevant in Pleistocene hominins (e.g. Kissel and Fuentes 2017, 2018). We do not argue that stone tools are necessary as mandatory indicators of such possibilities and lay out the *H. naledi *information in the context of the broader and increasing datasets and analyses for meaning-making behavior in Pleistocene hominins (see Figure 1 and table 1, and in the text).

Our point with this manuscript which we reiterate here is that “The increasing data for complex behavior and meaning-making across the Pleistocene should play a major element in structuring how we investigate, explain, and model the origins and patterns of hominin and human evolution” and we feel that the current evidence for H. naledi behavior contributes to the broader suites of data, hypotheses, analyses, and theory building in this endeavor.

The following is the authors’ response to the original reviews.

Before laying out how we addressed the specific comments on this manuscript we want to clarify the goal and intent of this paper to maximize effective critical reading of its contents. We appreciate and look forward to continued critique and enhanced discussion of this topic and argument.

Our starting point for constructing the argument in this manuscript is that H. naledi engaged in mortuary behavior. This emerges from the totality of the currently available data and analyses for Homo naledi in the Rising Star cave system, which support agential and intentional action by Homo naledi individuals in the transport of bodies to the Lesedi Chamber and Dinaledi Subsystem. We do feel that the data support the cultural burial hypothesis as well as the likelihood that at least some of the markings reported as engravings are non-naturally occurring (see Martinón-Torres et al. 2024) and made by Homo naledi. But these two elements are not necessary for the validity of the argument we pursue in this manuscript.

Our second key point is that gross brain size does not necessarily correlate with particular patterns of complex behavior in Pleistocene hominins. On this there is wide agreement, yet both scholarly and public arguments for the success of the genus Homo and the success of Homo sapiens have incorporated an assumption of a Rubicon of cerebral size. From this we propose a third point: that smaller brained Pleistocene hominins, including Homo naledi, are part of a Pleistocene hominin niche that includes patterns of complex social and cognitive behavior. Such behavior was historically considered to be exclusive to Homo sapiens but is now documented to occur earlier, across a range of hominin taxa in the latter half of the Pleistocene. We offer the case of H. naledi behavior in the Rising Star system as an example of this. This case contributes to the development of a broader approach to the cognitive, physiological, and behavioral framings of, and explanations for, Pleistocene hominin behavior.

Responses to specific critiques in the eLife reviews centered on this manuscript:

Reviewer #1:

All inferences regarding hominin behaviour and biology of Homo naledi, discussed by Fuentes and colleagues, are wholly dependent on the evidence presented in the archaeology preprints being true.

Reviewer #2:

Fuentes et al. provide a detailed and thoughtful commentary on the evolutionary and behavioral implications of complex behaviors associated with a small-brained hominin, Homo naledi…..While the review by Fuentes et al. highlights important assumptions about the relationship between hominin brain size, cognition, and complex behaviors, the evidence presented by Berger et al. 2023a,b does not support the claim that Homo naledi engaged in burial practices or symbolic expression through wall engravings.

Reviewer #3:

This paper presents the cognitive implications of claims made in two accompanying papers (Berger et al. 2023a, 2023b) about the creation of rock engravings, the intentional disposal of the dead, and fire use by Homo naledi. The importance of the paper, therefore, relies on the validity of the claims for the presence of socio-culturally complex and cognitively demanding behaviors that are presented in the associated papers. Given the archaeological, hominin, and taphonomic analyses in the associated papers are not adequate to enable the exceptional claims for nalediassociated complex behaviors, the inferences made in this paper are currently inadequate and incomplete.

We have clarified in the manuscript text and above why we argue that the inferences we are setting as core to our argument do not require cultural burial or engravings by H. naledi be demonstrated. However, we do clarify in the revision that the current evidence for the transport of dead conspecifics into difficult to reach areas deep into the cave system by naledi is well supported by the archeological and paleoanthropological data currently available (e.g. Berger et al. 2024, Elliott et al. 2021, Robbins et al. 2021, Hawks et al. 2017) and that this is the basis for our argument.

Reviewer #3:

The claimed behaviors are widely recognized as complex and even quintessential to Homo sapiens. The implications of their unequivocal association with such a small-brained Middle Pleistocene hominin are thus far reaching. Accordingly, the main thrust of the paper is to highlight that greater cognition and complex socio-cultural behaviors were not necessarily associated with a positively encephalized brain. This argument begs the obvious question of whether absolute brain size and/or encephalization quotient (i.e., the actual brain volume of a given species relative the expected brain size for a species of the same average body size) can measure cognitive capacity and the complexity of socio-cultural behaviors among late Middle Pleistocene hominins….Claims for a positive correlation between absolute and/or relative brain size and cognitive ability are not common in discussions surrounding the evolution of Middle- and Late Pleistocene hominin behavior.

We assert that claims for a positive correlation between absolute and/or relative brain size and cognitive ability are central—either explicitly or implicitly—in most arguments concerning cognitively complex behavior in the genus Homo. This is especially true for ideas about success of Pleistocene Homo relative to other hominins. We clarify this in the text offering various citations in support of this position (e.g. Meneganzin and Currie 2022, Galway-Witham, Cole, and Stringer 2019, DeCasien, Barton, and Higham 2022, Dunbar 2003, Kissel and Fuentes 2021, Muthukrishna et al. 2018, Püschelet al. 2021, Tattersall 2023).

Reviewer #3:

Currently, the bulk of the evidence for early complex technological and social behaviors derives from multiple sites across South Africa and postdates the emergence of H. sapiens by more than 100,000 years. Such lag in the expression of complex technologies and behaviors within our species renders the brain size-implies-cognitive capacity argument moot. Instead, a rich body of research over the past several decades has focused on aspects related to sociocultural, environmental, and even the wiring of the brain in order to understand factors underlying the expression of the capacity for greater behavioral variability. In this regard, even if the claimed evidence for complex behaviors among the small-brained naledi populations proves valid, the exploration of the specific/potential socio-cultural, neuro-structural, ecological and other factors will be more informative than the emphasis on absolute/relative brain size.”

While not at all denying the critically important and rich record of cultural complexity in the Late Pleistocene South African archeological record, we disagree that “the bulk of the evidence for early complex technological and social behaviors derives from multiple sites across South Africa and postdates the emergence of H. sapiens by more than 100,000 years”. We offer a range of examples and citations in support of our assertion in the text (esp. in pp12-14 and Table 1 and Figure 1)

We lay out the currently available data for such cultural complexity in Figure 1 with extensive documentation and citations for each case in the Supplementary material (both aa a table and a bibliography). We wholly agree with Reviewer 3 that “the exploration of the specific/potential socio-cultural, neuro-structural, ecological and other factors will be more informative than the emphasis on absolute/relative brain size” and are attempting to do just that in the manuscript.

Reviewer #3:

The paper presents as supporting evidence previous claims for the appearance of similar complex behaviors predating the emergence of our species, H. sapiens, although it does acknowledge their controversial nature. It then uses the current claims for the association of such behaviors with H. naledi as decisive. Given the inadequate analyses in the accompanying papers and the lack of evidence for stone tools in the naledi sites, the present claims for the expression of culturally and symbolically mediated behaviors by this small-brained hominin must be adequately established.

We respond to the first part of this critique above (regarding the other papers). But again, we emphasize that although we do feel that the argument for cultural burial is supported (see Berger et al. 2024 preprint) what we are arguing for in this paper is that the agential and intentional transportation of dead (mortuary behavior) is the sufficient factor undergirding our proposal. We do not agree that absence of recognizable stone tools at the site negates our proposal and assert that the context provided by Figure 1, and the data in the table for figure 1 in the SOM, in concert with the supported mortuary behavior (transport and emplacement of the dead) offer sufficient support for the argument we make in the text regarding brain size and the role of emotional cognition and complex behavior in the Pleistocene hominin niche and H. naledi’s participation in it.

Version 2

Abstract

Abstract

Explorations in the Dinaledi Subsystem of the Rising Star cave system have yielded some of the earliest evidence of a mortuary practice in hominins. Because the evidence is attributable to the small-brained Homo naledi, these analyses call into question several assumptions about behavioral and cognitive evolution in Pleistocene hominins. The evidence from the Dinaledi Subsystem, and at other locations across the Rising Star cave system may widen the phylogenetic breadth of mortuary, and possibly funerary, behaviors. These discoveries may also associate the creation of meaning making and increased behavioral complexity with a small-brained hominin species, challenging certain assertions about the role of encephalization and cognition in hominin and human evolution. We suggest that the hominin socio-cognitive niche is more diverse than previously thought. If true, technological, meaning-making activities, and cognitive advances in human evolution are not associated solely with the evolution of larger brained members of the genus Homo.

Main text

Contemporary humans engage in shared meaning via vocal, visual, tactile, and scent communication, often involving patterned use of bodies, objects, and materials. Of the very broad range of behaviors that create embedded shared meaning, the only ones that persist in archaeological contexts are those that leave material traces. Examples include aspects of mortuary behavior, engravings or paintings, personal ornaments, and use of ochre or other pigments. Prior to this century, most researchers readily accepted that such material traces could be interpreted as products of meaning-making behaviors when associated with Homo sapiens. But when such traces were associated with Neanderthals or other members of the genus Homo it was controversial. Previously, some archaeologists argued that the spectrum of behaviors in contemporary humans that involve shared meaning emerged as an integrated package in Late Pleistocene Africa, likely related to the dispersal of “modern” Homo sapiens throughout the world (e.g., Klein 1995). This view gave way to a recognition that meaning-laden material was used across a wider span of time, at least across part of the Middle Pleistocene (McBrearty and Brooks 2000), and not only by recent humans. In the last two decades, substantial evidence emerged of the extent of material evidence of meaning-laden behavior attributed to Neanderthals and other members of the genus Homo (Kissel and Fuentes 2018, 2021; Meneganzin and Killin 2024).

Currently a broad set of data demonstrates that some of these complex behaviors that involve shared meaning were manifested by multiple species and populations of the genus Homo including Homo heidelbergensis, Homo erectus, and possibly others (Galway-Witham, Cole, and Stringer 2019, Kissel and Fuentes 2018, 2021; Scerri and Will 2023) (Figure 1). Engravings of shell, bone, or rock surfaces have been identified in Middle Pleistocene or earlier contexts far from a range of African Homo populations. Some are also likely associated with Homo erectus (Joordens et al. 2014) as well as hominin populations that may have predated early Neanderthals in what is now Europe (Mania and Mania 1988; Sirakov et al. 2010). Evidence of ochre use occurs in archeological contexts across Africa and the Levant prior to 350,000 years ago, a time when H. sapiens has not yet been identified there (Ronen et al. 1998; Watts, Chazan, and Wilkins 2016; Dapschauskas et al. 2022). The control of fire by hominins is demonstrated in Early and Middle Pleistocene contexts where most researchers accept that H. erectus was present (Brain and Sillen 1988; Alperson-Afil 2008; Goren-Inbar et al. 2004; Hlubik et al. 2019; MacDonald et al. 2021). Mortuary evidences are claimed in association with hominins that predate or are not H. sapiens (Carbonell and Mosquera 2006). These geographically and temporally varied instances could be the result of taphonomic or dating issues (Püschel et al. 2021), but it is likely, given the increasing diverse temporal and geographic discovery of these behaviors and material that such complex behaviors associated with shared meaning were manifested by multiple populations/species of the genus Homo in addition to Homo sapiens. But how important was brain size to the evolution of these behaviors?

Archaeological evidence of culturally-mediated, meaning-making, behaviors.

Dots represent different sites and the error bars are the maximum and minimum dates when available. See Table for details and See supplemental material for references.

There is substantive evidence that approximately 250-350,000 years ago Homo naledi, a small-brained hominin, transported deceased conspecifics into difficult to access locations in the Rising Star Cave system in what would, in humans, be described as a mortuary behavior (Berger et al. 2024). The use of deep areas of the Rising Star cave system for these behaviors implies considerable social collaboration, coordination, and planning. In the context of the subterranean Dinaledi Subsystem, these activities likely also required a light source, again implicating a depth of planning and coordination. What stands out as a possible contradiction is that Homo naledi fossil crania are small. With endocranial volumes ranging between 450 ml and 610 ml, this species overlaps in brain size with australopithecines, having smaller brains on average than Homo erectus and much smaller than modern humans or Neanderthals (See Figure 2 and section Reconsidering Brain Size).

Endocranial volume estimates for hominin cranium.

Error bars represent the maximum and minimum ages for specimens when available. See supplemental material for references. Hawks, John. 2023. Endocranial volumes for fossil hominins (dataset). Figshare https://doi.org/10.6084/m9.figshare.22743980

The information from H. naledi cannot be considered in isolation; it joins the broad array of data for meaning-making in Pleistocene hominins (Kissel and Fuentes 2018, 2021; Malfouris 2013). These behaviors reflect social groups that maintained solidarity, social coordination and cooperation in a mode not evident in living great apes but characteristic of contemporary humans. Here we offer an analysis of the reported complex behavior in the small-brained Homo naledi and suggest a suite of implications this has for our understanding of the relationships between brain size, cognition, complex behavior and the evolution of the genus Homo across the Pleistocene. These implications also query the driving forces behind encephalization and its relationship to the emergence of complex behaviors in hominins and other animals (see Tatersall 2023).

Mortuary behavior as meaning making

Mortuary behavior has been defined as actions by individuals relating to the death of other individuals. Many kinds of non-human animals have been observed to engage in mortuary behavior upon the death of another individual in the same social group (Piel and Stewart 2015; Gonçalves and Carvalho 2019; Pettitt and Anderson 2020). Mortuary behavior may include the manipulation, inspection, or movement of dead bodies or body parts, or modifying behavior in proximity to dead or dying individuals. Species which manifest strong emotional bonds between individuals, including many primates, elephants, and cetaceans, may continue to interact with a corpse for a period of time after the individual’s death. Many such examples involve mothers who continue to carry a dead infant, often for up to a week after death (Goodall 1977; Boesch and Boesch-Achermann 2000; Biro et al. 2010; Fasching et al. 2010), and in some cases other individuals such as unrelated males have been observed to carry or interact with dead infants (Merz 1978). Species with strong social bonds often exhibit emotional responses to dead or dying individuals, ranging from surprise and fear to prolonged grief (King 2013). Cannibalism is also a form of mortuary behavior observed in some non-human primates.

Scientific examination of mortuary behavior across living species has been limited. Piel and Stewart (2015) noted the publication bias related to mortuary behavior in non-human species. Field researchers may hesitate to publish observations that appear to be isolated cases or anecdotes, and as Piel and Stewart point out, almost no studies are negative reports that claim that no mortuary behavior occurs in a species. In recent years, a growth in interest in mortuary behavior in nonhuman primates and other social mammals has been fueled by attempts to understand ancestral hominin mortuary practices, including those evidenced in Neanderthals (Pettitt 2018; Pettitt and Anderson 2020; Pomeroy et al. 2020). Discussion of the evolutionary background of mortuary behavior in non-human species has focused upon possible adaptive or maladaptive consequences of the behavior (Piel and Stewart 2015). Less attention has been given to the proximate neural, psychological, or social mechanisms that give rise to mortuary behavior in various species.

There are clear differences between the mortuary behavior repertoire of humans and the mortuary behavior observed in many other living primates. In humans some mortuary behavior is ritualized, meaning that the behavior is systematized and repeated. Funerary behavior is a category of mortuary behavior which is defined as specific activities relating to the disposal of the dead and to their subsequent commemoration (Pettitt 2011; 2018; Pettitt and Anderson 2020). While funerary behavior is not characteristic of other primates or social mammals, human mortuary behaviors do include many patterns that are also seen in some non-human species, such as the temporary curation of corpses or body parts, disposal of corpses without commemoration, dietary or starvation cannibalism, and manipulation of corpses out of curiosity or fear (Engelke 2019; Silverman et al. 2021).

To the extent that human behaviors surrounding death are different from non-human primates, those differences reflect primarily two mechanisms. One dynamic underlying human mortuary behavior is the shared, learned cultural traditions concerning death. These traditions vary extensively across human cultures and may include an understanding of the permanence of death, religious beliefs and practices concerning death, and scientific and practical knowledge about the causes of death. When anthropologists examine funerary practices, they often describe this kind of cultural knowledge (Engelke 2019; Silverman et al. 2021). Another dynamic involved is an evolved emotional cognition including emotional self-awareness and regulation. In human cultures these two dynamics are interconnected: religious rituals help grieving individuals by providing social support for emotional regulation and for processing and resolving relational trauma, for example. Superficially similar instances of mortuary behavior in different cultures or species may sometimes involve different proximate mechanisms. For example, ritual and funerary behavior both involve the learned repetition of behavioral and emotional states (Pettit 2018, Silverman et al. 2021). In effect a shared capacity for culture integrates with complex emotional cognition, involving a depth of emotional bonds, capacities for emotional commitments (which necessarily come alongside grief at loss, and a need to process relational trauma) and complex emotional regulation. Both have been practiced by people with entirely different or even incommensurable cultural traditions concerning death, which differ widely, yet fulfil similar emotional needs. Hence the interpretation of physical evidence for mortuary behavior should consider the range of cultural and cognitive mechanisms that may be at play and how they interact, which may give rise to different equally plausible explanations for the pattern of evidence.

Rising Star evidence and context

The findings from the Rising Star system strongly support a scenario where members of the H. naledi community carried the bodies of dead conspecifics to more than 30 meters below the surface, over more than 80 meters of underground passages in a difficult and dangerous subterranean environment (Berger et al. 2024; Elliott et al. 2021). The available evidence demonstrates some aspects of mortuary behavior manifested by H. naledi. Multiple lines of material evidence show that corpses were manipulated, both at the time of death and afterwards. The distribution of skeletal parts across the Dinaledi Subsystem could not have arisen from deposition at a single point of entry to the subsystem with gravity-driven movement of bodies or bones (Berger et al. 2024). Both the spatial arrangements of skeletal material and the form and composition of sediments rule out water flow or mud flow as mechanisms for transport (Dirks et al. 2015; Wiersma et al. 2020; Brophy et al. 2021; Berger et al. 2024). Movement of remains by Homo naledi is the best hypothesis for the emplacement of bodies or remains. The rapid emplacement of some bodies into sediment prior to decomposition and continued support by sediment through the process of decomposition, together with evidence of disruption of surrounding sediment layering, all suggest that some bodies were interred within shallow holes and covered before soft tissue decomposition occurred (Berger et al. 2024). Some bodies were manipulated after deposition, as evidenced by the selective reworking of the Puzzle Box area of the Dinaledi Chamber, leading to fragmentation of some skeletal elements, disaggregation of body parts, and commingling of elements from different individuals (Berger et al. 2024). The presence of young children and infants within the sample likely also reflect the manipulation of bodies by other, presumably older, individuals (Berger et al. 2024, Delezene et al. 2023).

Many aspects of the mortuary behavior represented by this evidence remain unclear. The evidence does not show whether postmortem manipulation in the Puzzle Box area was deliberate and commemorative in intent, or whether this manipulation was an incidental result of the introduction of subsequent bodies into the Puzzle Box are or other activities in the Dinaledi chamber. Archaeological collection and excavation of remains across the Dinaledi subsystem has revealed varied dispositions of different individuals. Some of the skeletal remains show no evidence of intentional disruption after deposition, while others underwent marked and selective reworking after initial deposition, and other skeletal parts were isolated in long fissure passages (summarized in Berger at al. 2024). It is unclear whether these differences reflect intentional differences in mortuary behavior, whether they reflect changing traditions over time, or whether they resulted from a lack of precise patterning of mortuary activity. The duration of mortuary activity in the cave system is not known, nor is it known whether the remains in the Lesedi Chamber and in the Dinaledi Subsystem represent activities of the same group, culturally related groups, or unrelated groups. The anatomical evidence also suggests that the sample of H. naledi individuals may be biased with overrepresentation of one sex (Delezene et al. 2024), and it is not known whether this bias was an intentional result of the mortuary behavior. While a majority of the remains associated with mortuary behavior of H. naledi occur within deep cave areas, it is not clear whether this behavior was limited to these spaces or whether the observed evidence may represent only a small part of a more extensive pattern.

The use of deep cave spaces as part of the mortuary behavior of H. naledi provides additional evidence about the social and emotional mechanisms of this species. The subterranean environment used by H. naledi is physically challenging for today’s researchers. A longstanding question is whether the system was equally challenging for H. naledi. The journey to the Dinaledi subsystem from any known or reasonably hypothesized incursion point involved strenuous scaling and navigation of complex three-dimensional topography across distance, multiple chambers, passages, climbs and descents (Elliott et al. 2021, Robbins et al. 2021, Berger et al. 2024:SOM). To accomplish this H. naledi had to coordinate their behavior and collaborate to move the bodies to a specific location inside the Rising Star cave system. Several aspects of the biology of H. naledi suggest that this species may have been better at underground movement than today’s humans: Adult Homo naledi individuals had smaller body size than even small-bodied caving team members today (Garvin et al, 2017), and a body plan that was more suited for climbing and passing through narrow and restricted cave passages (e.g. Kivell et al, 2015; Feuerriegel et al, 2017, 2019; Williams et al, 2017; Traynor et al, 2022). Still, movement into these spaces would have had high energetic costs and carried some risk for the H. naledi individuals undertaking the behavior. Doing so while carrying a corpse would have entailed additional energetic costs. Given the structural complexity of the cave system layout (Berger et al. 2024, Elliott et al. 2021, Robbins et al. 2021), there must have been some form of explicit communication (tactile, vocal, and likely visual) for coordination of movement and actions, and the potential use of fire as a light source, between the H. naledi undertaking the behavior. Such coordination and specific set of actions around the treatment of deceased conspecifics is more methodologically extensive, energetically costly, with higher risk of injury than any reported for other primates and non-human animals to date (King 2013). This behavior is also more complex and multifactorial than that reported for the one earlier case of hominin mortuary behavior (Sima de Los Huesos, Carbonell and Mosquera 2006). There are no clear direct fitness benefits nor any indication of particular proximate functional stimuli for this suite of behaviors.

The subterranean environment used by H. naledi is not only physically challenging but is also emotionally and physiologically challenging, reflecting a particular engagement with difficult underground spaces not common in the archaeological record of that time. Dark enclosed spaces, where visual perception is curtailed, can create a state of emotional arousal profoundly affecting perceptual, cognitive, physiological, and social systems (Zuccarelli et al. 2019; Kedar et al. 2021), even with some form of illumination. It is very likely that H. naledi used forms of tactile communication and a range of proprioceptive tactics to navigate and communicate in the Rising Star cave system (e.g Dunbar 2010, Waller and Hodgson 2013). However, while such modes of social interactions can enable a range of coordination, it is likely that some form of illumination as also necessary to undertake the behavioral patterns reported (e.g. Pettitt, 2022).

In humans, and other diurnal primates, sensory deprivation through reduced or a lack of consistent visual clues creates a heightened sensitivity to other senses (and thus augments proprioceptive dynamics) as well as prompting experiences of visual disturbances, hallucinations, and disorientation (Hodgson 2021). Experiences of these types of extreme and unusual environments, though often inducing fear responses, can also facilitate powerful bonding experiences (Steidle, Hanke, and Werth 2013). Furthermore, the interoceptive nature of bodily awareness such as that experienced during traversing a complex cave system enhances empathy (Ernst et al. 2013) which may be augmented and deployed in tactile and vocal communication and coordination. This range of substantive emotional, psychological, and physiological reactions may explain why experiences in deep dark caves are often associated with a sense of the transcendent in contemporary humans (Montello and Moyes 2012, Kedar et al. 2021) and given the broad range of sensory commonalities across diurnal anthropoids, and especially apes, such experiences likely had comparable impacts on H. naledi and other Pleistocene hominins using subterranean spaces. We argue that careful and coordinated treatment of the dead on several occasions, in these subterranean environments, implies particularly strong social and emotional bonds and some shared understanding of meaning (Pettitt and Anderson 2020) in the handing of the dead by H. naledi.

As with most examples of Pleistocene mortuary behavior, researchers should be cautious when comparing to modern analogs. Mortuary and funerary behaviors in the past need not map directly to the practices of contemporary or Late Pleistocene humans (see Berger et al. 2024). As recently suggested by Meneganzin and Killin (2024: 26) in a discussion of Neanderthal aesthetics: “…we should not be surprised if the search for Neanderthal aesthetic practices suggestive of an aesthetic sense requires taking a different route, at least sometimes, to the search for (paradigmatic examples of) human aesthetic practices.” This suggestion is likely applicable to a range of Pleistocene hominin behavior. An important point of comparison is the Neanderthal use of deep caves, in certain cases reflecting a substantial duration of activity and repeated use (Jaubert et al. 2016; Baquedano et al. 2023).

That this high-risk, high-cost, no-overt-direct-fitness-benefit behavior was undertaken repeatedly by multiple members of an H. naledi community indicates a valued, likely cultural, tradition with a social and emotional function. The combination of features in the behavior and the context in which it was undertaken (in deep caves with the likely use of fire for illumination), suggests a level of cognitive/semiotic meaning-making capacity in H. naledi (e.g., Kissel and Fuentes 2017; 2018; 2021) that matches some similar assessments of other populations of the genus Homo during earlier, the same, and later time periods (Figure 1). It is our hypothesis that the repetition of mortuary activities within the Rising Star cave system reflects a form of shared memorialization. This hypothesis emerges from the fact that shared attention and joint action were necessary to generate the evidence within the system, and these actions were repeated over some period of time. The hypothesis of shared memorialization does not depend upon individual interments being repeated with identical steps. What supports the hypothesis is the repeated pattern of cave use and the context and distribution of the remains. The collective practice by H. naledi, coupled with social and emotional investment, helped transform the ‘space’ of the Dinaledi Chamber and Hill Antechamber to ‘place’ (Low and Lawrence-Zúñiga 2003) through the pattern of mortuary and possible funerary behavior (e.g. Silverman 2008). The coordinated treatment of the dead on several occasions, in these subterranean environments, implies particularly strong social and emotional bonds and some shared understanding of meaning (Pettitt and Anderson 2020) in the handing of the dead by H. naledi.

Some form of mortuary behavior by H. naledi is a supported hypothesis (Berger et al. 2024). However, there is substantive criticism of the assertions of cultural burials in the Dinaledi subsystem (see Berger et al. 2024, Martinon-Torres et al. 2023, Foecke et al. 2024). Recently Pettitt (2022) laid out the three key criteria for assessing whether or not the Homo naledi remains in the Dinaledi subsystem represent actual funerary behavior: a) is there an as-yet unmapped entrance into the Dinaledi Chamber? b) Is there any evidence of artificial lighting in the cave system, and c) Is there evidence that it was dead bodies, rather than body parts that were carried into the chamber? The first query has been repeatedly addressed and no other options for alternative ingress into the Dinaledi Subsystem involving movement across less than 80 meters of structurally complex crevices, chambers, and passages involving a descent to a depth of ∼30 meters below the surface have been found or potentially identified (Elliott et al. 2021, Robbins et al. 2021; Berger et al. 2024). There is reported evidence of fire use (hearths and charcoal and smoke scarring of surfaces) in the Dinaledi system (see Bower 2022, but also see Martinon-Torres et al. 2023) although the age of these occurrences has not been reported yet, so its association with H. naledi is currently correlational. However, the correlational support is a reasonable hypothesis given there is no evidence of any kind across the last decade of investigation of any habitual activity by ancient or recent Homo sapiens in the Dinaledi Subsystem or even deep proximal spaces such as the Dragon’s Back Chamber or Lesedi Chamber. Articulated skeletal elements do make up a substantive percentage of the H. naledi remains reported in the Dinaledi system, especially in feature 1 and in the Hill Antechamber feature (Berger et al. 2024). While the one other major locus of remains in Dinaledi (the Puzzle Box feature and surface collections) have a majority non-articulated materials, there are articulated remains present and there is strong evidence that disarticulation resulted from postdepositional reworking, likely by H. naledi (Berger et al. 2024). Therefore, we suggest that the currently available evidence tentatively meets Pettit’s (2022) criterion for funerary action. As it stands, the whole of the evidence supports the hypothesis that the H. naledi remains in the Dinaledi subsystem are one of the two earliest examples of a mortuary practice in a hominin, and potentially offer the earliest evidence of multiple interments, postdepositional reworking, and thus funerary actions by a hominin.

In addition to the above, the locations, contexts, and the inferred behavior associated with the H. naledi remains most likely demonstrate shared meaning-making activity (Kissel and Fuentes 2017, 2018). Certainly, if the remains are postdepositionally re-worked, and/or do represent burials, and one accepts the correlational association, and validity, of engravings near the interment sites with H. naledi (e.g. Berger et al. 2023) they do. But even if one only accepts the transport to and placement of bodies in the Dinaledi Chamber and Hill Antechamber locations (and the Lesedi Chamber, Hawks et al. 2017, Berger et al. 2024) there remains a robust argument for mortuary behavior and the assignation of shared meaning to it. Most documented mortuary and funerary practices have been attributed to Homo sapiens and Neanderthals, and aside from the Sima de los Huesos site (Carbonell and Mosquera 2006) and Dinaledi, most such evidence is later the Pleistocene (Figure 1 and Table 1). Evidence of funerary behavior is generally assumed to require human-like cognitive capability (Pettitt 2018). If such behavior is indeed present in a small-brained hominin it suggests that increases in brain size/EQ are not a necessary precursor for the appearance of complex meaning-making behavior in hominins.

A role for emotional cognition?

The achievement of social collaboration and social solidarity in humans relies upon emotional cognition and emotional regulation. A broad array data supports the hypothesis that emotional regulation and self-awareness were prerequisites of human social behaviors involving solidarity and cooperation, including cultural learning, language, and provision of extended care (Spikins et al. 2018).

Humans share many building blocks of emotional cognition with other mammals, and some complex abilities with other primates. We share the same visceromotor and sensorimotor foundation for emotions with other mammals for example (Steklis and Lane 2013). Moreover, a range of common emotional responses in humans have now been documented in other primates, particularly in apes, through measurement of heart rate and skin conductance as well as more recently pupil mimicry and infrared thermography (Nieuwburg, Ploeger, and Kret 2021). Interpersonal emotional interactions have a common basis. Emotional contagion is apparent in monkeys and apes, and apes in particular demonstrate a level of empathy though yawning and even sympathy through active consolation (Romero, Castellanos, and de Waal 2010; Preston and de Waal 2002). Diverse primate species have the cognitive ability to infer emotional meaning from expressions (Nieuwburg, Ploeger, and Kret 2021). Moreover, there is anecdotal evidence for the foundations of cognitive empathy in targeted helping within apes (Koski and Sterck 2010).

One important aspect of human emotional cognition is the ability to regulate emotions by bringing feelings into “rational” thought (Green and Spikins 2020). Humans communicate and engage in shared intentions, and meaning-making, to a degree not seen in other animals, and demonstrate motivations to share emotions, experiences and activities with other persons (Steklis and Lane 2013; Fuentes 2018). Emotional self-awareness, and with it the capacity to regulate emotions—that is, to calmly tolerate difficult feelings and bring them into rational thought—is key to many human social behaviors. Emotional self-awareness is essential to translating empathy into systematic compassionate helping for others for example (Spikins 2022). Furthermore, emotional self-awareness allows the regulation of interpersonal emotional vulnerabilities which foster connection and collaboration such as fearfulness (Grossmann 2022). Self-conscious emotions also more broadly regulate social behaviors in general (Beer et al. 2003) and conversely impaired emotional awareness interferes with normal social function in both clinical and nonclinical populations (Gu et al. 2013). Perhaps most importantly emotional self-awareness provides the basis for emotional commitments which bring high levels of give and take to social relationships. Capacities to make emotional commitments to group interests are often demonstrated through costly signaling in risky, nonfunctional ways (Hall et al. 2015; Lang and Kundt 2023).

Another important aspect of human emotional cognition is the presence of emotional cues that are absent or not well developed in other living species. The depth of human emotional commitments come with costs. Firstly, living humans accentuate emotions associated with social control, including shame and guilt, which are not manifested in similar ways in other great apes (Boehm 2012; Turner 2014). Sympathy for others and guilt both emerge early in human ontogeny, with sympathy providing a basis for prosocial orientation in very young children and guilt helping provide motivation for repair of social ruptures (Vaish and Grossmann 2022). These have sometimes been examined within the framework of moral emotions, which would additionally include emotions such as contempt, gratitude, and disgust (Fitouchi et al. 2023). These aspects of emotional cognition in humans provide a foundation for prosociality and social solidarity. Both emotional self-awareness and moral emotions function to regulate social interactions and maintain cohesive and cooperative social relationships. Secondly, emotional bonds with high degrees of give and take, and emotional commitments, which drive behaviors such as risky hunting or caring behaviors necessarily involve relational trauma at loss, regulated and resolved through social relationships and cultural practices.

The H. naledi evidence suggests that a human-style conscious emotional awareness was present in this hominin despite its small brain size. The hominins carrying out these mortuary activities would need to be able to bring their emotions into “rational thought” in order to both be aware of their own grief and communicate and coordinate shared intentions over the bodies of the deceased. Moreover, they would need to be able regulate their emotions (hold feelings in calm awareness) such that they were able to mutually engage in coordination to carefully negotiate extensive, complex subterranean landscapes, despite the risk and complexity of such behavior, to transport bodies into the Dinaledi Subsystem. This level of emotional regulation, coordination, and awareness is markedly different from the generally personal (such as corpse interaction), sometimes numb (corpse avoidance) and often disordered displays of grief in our nearest relatives (Pettitt and Anderson 2019). The shared and planned transportation and placement of several bodies in the Rising Star system is also evidence of a shared sets of cognitive commitments, beliefs, or assumptions about meaning and action, something similar to what one would term “shared grief” and/or “shared belief” in contemporary humans. The behavioral sequences required for mortuary action also suggests a form of shared memorialization, or at least more behaviorally and communicationally complicated shared attention and action to achieve the deposition of the bodies in the locations in the Rising Star system. Regardless of whether one accepts the interpretation of burials and the presence and association of engravings with H. naledi, the underlying cognitive processes associated with just the transport and placement of H. naledi into the Dinaledi subsystem indicates a level of conscious emotional awareness that enables and is associated with extensive shared intentionality, forward planning, and repeated cultural behavior involving bodily risk. Equally complex use of caves by Neanderthals (Jaubert et al. 2016; Baquedano et al. 2023) demonstrate a similar emotional self-awareness, and production of highly symmetrical stone tools is also potentially indicative of certain aspects of emotional awareness and regulation in earlier member of the genus Homo (Green and Spikins 2020). Furthermore, this evidence suggests a depth of emotional commitments, with a willingness to take risks and costs on another’s behalf. Social understanding of emotions is widely accepted as adaptive in an evolutionary context (Nieuwburg, Ploeger, and Kret 2021) and emotional awareness is associated with better life outcomes in contemporary human contexts (Smith et al. 2023).

That complex emotional cognition is not unique to Homo sapiens should not be surprising. But it is not strictly associated with overall brain size. The fact that a small-brained hominin displays these sorts of behaviors suggests that the neurological capacity enabled by a larger than 1000cc brain cannot be the only factor, or necessarily the main factor, enabling the kind of emotional cognition that is considered a central factor in human evolutionary success. Particular brain areas are related to emotional regulation including anterior cingulate cortex (Giuliani et al. 2011) and amygdala (Davidson et al. 2007). Much in the prefrontal cortex has also been implicated in emotional regulation and executive control (Davidson et al. 2007). Recent approaches also focus on large-scale brain networks being implicated in emotional regulation (Morawetz et al. 2020; Pozzi et al. 2021; Rieck et al. 2024). The associations between brain areas or networks and emotional regulation and self-awareness are studied by considering how these cognitive traits correlate with the volumes of various brain areas, and by the activation of brain areas as indicated by oxygen consumption during experimental tasks. Neither kind of study can be compared easily with studies of fossil hominins. The only data on brain anatomy from fossil hominins are the volume of the endocranial cavity and the few sulcal and gyral patterns that imprint on the endocranial surface. These data provide one suggestive indication that emotional regulation may have been important across Homo, which is that species of the genus have humanlike frontal cortex configurations, including Homo floresiensis and Homo naledi (Falk et al. 2005; Holloway et al. 2018; Hurst et al. 2024).

The anatomical data alone do not answer when humanlike emotional self-awareness and emotional regulation first evolved. Some have suggested that hominins would have been under selection for these traits early in their evolutionary history because of the need for cooperating, cohesive groups in open habitats with high predation (Turner 2014). Others have suggested that such behavior is a hallmark of the genus Homo, or that abilities such as technical learning or the routine use of fire (Twomey 2019) could only have been manifested in species with emotional cognition that was similar in ways to recent humans. Spikins and Green (2020) specifically pointed to several aspects of Early Stone Age artifacts like Acheulean handaxes as possible indicators of self-control including inhibition and conscious regulation of emotions. The reported H. naledi behavioral activities, may have depended on emotional commitments to others combined with a set of cultural beliefs/practices, a high level of emotional awareness to manage these, and in turn collaboration with extensive coordination.

Homo naledi behavior in broader perspective

The behavior patterns manifested in the Rising Star cave system have a distinctive place within a broader global context of mortuary behavior. The repeated mortuary behavior involving more than 30 individuals from Sima de los Huesos, Spain, is substantially earlier than evidence from the Dinaledi Subsystem (Carbonell and Mosquera 2006; Arsuaga et al. 2014; Vives 2015; Sala et al. 2024). Cutmarks on hominin individuals such as the Bodo 1 skull from Bodo, Ethiopia, and the StW 53 skull from Sterkfontein, South Africa, may result from even earlier mortuary activity by hominins (White 1986; Pickering et al. 2000). In the case of StW 53, the evidence is associated with a skull that many researchers attribute to Australopithecus, although the claim of hominin-produced cutmarks has been disputed (Hanon et al. 2018). Curation of hominin skeletal remains was part of mortuary behavior at Herto, Ethiopia before 160,000 years ago (White et al. 2003; Clark et al. 2003). Archaeological evidence of mortuary behavior becomes increasingly common in later contexts, and within the Late Pleistocene, burials and other kinds of funerary behavior have been attributed to both modern humans and Neanderthals. Within this broader pattern, the Rising Star evidence stands out in two ways: the early possible occurrence of burials (Berger et al. 2024; Dirks et al. 2017), and the position of H. naledi as a phylogenetic outgroup when compared to modern humans and Neanderthals (Dembo et al. 2016; Argue et al. 2017; Caparros and Prat 2021).

Mortuary behavior is only one category within a broader set of behavior patterns that emerge from shared meaning within social groups (Table 1 and Figure 1). Another category of evidence is the patterned engraving of lines on bones, shells, rocks, or rock walls. Providing geochronological context for such engraved features on rock walls is challenging as it requires that walls themselves be buried in sediment or that engraved features be partially obscured by material susceptible to dating, such as calcite crusts. The engraved lines and percussion marks observed within the Dinaledi Subsystem (Berger et al. 2023) are not an exception to this challenge; however, the widespread presence of H. naledi remains within this space and absence of any evidence of modern human activity other than entry by recent explorers, make it a reasonable hypothesis that H. naledi individuals were authors of these marks. Such a hypothesis is reasonable when considered in a global context. Engraved bones are known from several sites as old or older than the Dinaledi Subsystem, including marked bones, ivory, and stone from Bilzingsleben, Germany (Mania and Mania 1988), an engraved shell from Trinil, Indonesia (Joordens et al. 2016), and an engraved bone from Kozarnika, Bulgaria, once suggested to be around 1.4 million years old (Guadelli and Guadelli 2004), although recent geochronological work suggests that the layer is likely early Middle Pleistocene in age (Heydari et al. 2022). A larger array of such evidence is known from later Middle Pleistocene and Late Pleistocene, including examples of engraved hominin bone.

Evidence of the entry into caves by hominins occurs across the span of the fossil record of South Africa, three million years or earlier. What is apparently somewhat unusual in the Rising Star Cave is the repeated hominin use of deep caves, where illumination is necessary, including nonutilitarian use of the space. No evidence of such behavior has yet been found earlier than the Dinaledi Subsystem. Still, the evidence for manipulation of stalagmites and built structures within Bruniquel Cave, France (Jaubert et al. 2016), presumed to have been done by Neanderthals, as well as later Neanderthal marking within deep caves such as Cueva de Ardales, Spain (Pitarch Martí et al. 2021), or areas far from cave entrances such as at Gorham’s Cave, Gibraltar (Rodríguez-Vidal et al. 2014), bring to mind the activity seen in the Rising Star cave system. Within South Africa, the site of Wonderwerk Cave has evidence that may be contemporary or slightly later than the Dinaledi Subsystem including ochre and quartz crystals taken approximately 100 m from the cave entrance, although within this linear and large cave the entrance is always visible from the excavated area (Chazan et al. 2020).

The sustained use of deep cave areas is, in humans, tied to the use of fire for illumination. Hominin control of fire in South Africa long predates the Rising Star evidence, with controlled fire indicated in Member 3 at Swartkrans (only ∼ 800m from the Rising Star Cave), and in Early Stone Age context at Wonderwerk Cave, by one million years ago (Brain and Sillen, 1988; Brain, 1993; Berna et al., 2012). Both Paranthropus and early Homo were present in South Africa during that period and occur in association with combustion evidence (Brain and Sillen, 1988; Brain, 1993). Additionally, cave use by hominins across multiple locations during the period of H. naledi and earlier (e.g. Wonderwerk in South Africa and Bruniquel in France) also involved use of fire (Berna et al. 2012, Jaubert et al. 2016). While we cannot yet be certain of the exact modes, intensity, duration, and quality of the fires used by H. naledi in the Dinaledi subsystem, it is a strong assumption that they at least provided flickering and moderate intensity light sources.

Use of ochre and other pigments is another category of behavior linked to meaning-making (Dapschauskas et al. 2022; Kissel and Fuentes 2018). Material evidence of pigments carried or used by hominins has been found at several sites earlier than the Dinaledi Subsystem, including sites within Africa, as well as within southwest Asia and south Asia (Figure 1). Non-utilitarian objects that were transported by hominins into sites include some with physical or iconic resemblance to human figures, including objects from Berekhat Ram, Israel (Goren-Inbar 1986) and Tan-Tan, Morocco (Bednarik 2003), both from the later Middle Pleistocene.

This pattern of evidence shows that H. naledi and other populations of the genus Homo overlap temporally in the expression of meaning-making behavior. The material evidence indicates some degree of shared socioemotional and cognitive processes. Middle Pleistocene hominins varied in brain sizes and cranial and post-cranial morphologies, and many of these varied populations share increased evidence for meaning-making (Figure 1). Such behavior is neither “modern” nor exclusive to larger brained Homo sapiens (and Neanderthals). Whilst this adds further evidence to our understanding of the emergence of hominin cognition there are also wider evolutionary implications. Much like the evolution of social emotional abilities in other primates (Nieuwburg, Ploeger, and Kret 2021) the behavioral evidence for small-brained H. naledi may suggest that some degree of analogous, and homologous, evolution underlies social emotional complexity in humans.

Reconsidering brain size

Since the nineteenth century, ideas about the evolution of human behavior have tended to emphasize that larger brains evolved to enable more complex behavior. In the most general sense this is surely correct. Large brains are expensive to maintain and grow (Isler and van Schaik 2009). Considering these costs, the large brain sizes manifested in some hominin lineages would not have evolved if they were not reliably correlated with survival or reproduction. It is often assumed that a large brain was an essential step towards a uniquely human cognition, social relationships and culture (Dunbar 2003; Muthukrishna et al. 2018). Several hominin lineages during the Pleistocene did experience evolution of larger overall and relative brain size, as measured by endocranial volume (Figure 2) and body size. Nothing in the Pleistocene behavioral record refutes the basic idea that such lineage-specific increases in brain size evolved alongside some behavioral adaptations.

But the Pleistocene record today does not support the notion that every behavioral adaptation was mediated by overall brain size. Planning and forethought in stone tool production both preceded the first appearance of Homo (Harmand et al. 2015). Early populations of H. erectus dispersed from Africa into Eurasia before 1.8 million years ago, and the first substantial sample of these hominins has endocranial volume ranging from 550 ml to 730 ml (Ponce de Léon et al. 2022). The first bifacial tool traditions were developed before 1.76 million years ago, within a geographic and temporal context where the only extant groups of hominins had small average endocranial volumes (Lepre et al. 2011). The use of fire emerged in excess of a million years ago in Africa (Hlubik et al. 2019) at a time when the average endocranial volume among crania attributed to H. erectus was around 750 ml. It is credible that ancestors of H. naledi were among these early fire users. The phylogenetic arrangements of species within the genus Homo remain uncertain (Hawks et al. 2017; Argue et al. 2017; Dembo et al. 2016; Caparros and Prat 2022). Still, much evidence suggests that the lineages leading to hominin species with small brain size including H. naledi and H. floresiensis emerged from within this early Pleistocene evolutionary context (Dembo et al. 2016; Hawks and Berger 2019). These species would have been part of the hominin niche that also gave rise to other later Pleistocene members of the genus Homo (Mondanaro et al. 2020).

It is also important to be precise about what the small sample of fossils really tells us about differences in brain size between lineages (Figure 2). Known Homo naledi individuals for which endocranial volume estimates can be made (n = 5) have volumes ranging from 450 ml to 610 ml (Holloway et al. 2018; Hurst et al. 2024), the single known endocranial volume for H. floresiensis is 426 ml (Kubo et al. 2013; Falk et al. 2005). With so few fossils, these samples are unlikely to provide accurate understanding of the average of either species. Moreover, both samples may be biased. The LB1 individual of H. floresiensis is inferred to be female based on postcranial morphology (Brown et al. 2004) and male individuals would likely have had larger endocranial volumes. The low degree of variation in size within the H. naledi sample suggests that the sample may be biased toward representation of one sex (Delezene et al. 2024), again indicating that the sample average may misrepresent the species mean. Both samples have smaller average endocranial volume than H. erectus, yet many the fossils attributed to this species within Africa have estimated endocranial volumes within or just above the range observed in H. naledi, including DNH 134 (Herries et al. 2020), KNM-ER 42700 (Neubauer et al. 2018), DAN5/P1 (Semaw et al. 2020), KNM-OL 45500 (Potts et al. 2004), and OH 12. While global H. erectus endocranial volume increased over time (Leigh 1992), fossils attributed to this species with small endocranial volumes occur across the entire temporal range of the species within Africa.

Different hominin lineages manifested differences in brain organization, and these do not always correspond to changes in brain size (Holloway et al. 2018; Hurst et al. 2024; Ponce de Léon et al. 2021). Endocast evidence from across the genus Homo suggests that humanlike frontal cortex development and morphology occurs across a broader array of species than large brain size (Falk et al. 2005; Holloway et al. 2018; Ponce de Léon et al. 2021). Homo floresiensis and Homo naledi both share some humanlike aspects of frontal lobe organization while they both have brain sizes within the range of Australopithecus (Falk et al. 2005; Holloway et al. 2018; Hurst et al. 2024). Early Homo erectus is argued to be polymorphic in frontal lobe organization (Ponce de Léon et al. 2021), and as noted above many fossil crania attributed to this species have endocranial volumes smaller than 650 ml (Lordkipanidze et al. 2013; Ponce de Léon et al. 2021; Herries et al. 2021; Semaw et al. 2020).

Looking forward

The initial rise of complex social behaviors in hominins may have been fueled by evolution of emotional cognition alongside other cognitive processes. Meaning-making by hominins includes a variety of categories of behavior that are joined together by the role of shared intention, repetition, and social collaboration. Mortuary behavior is one of these. Varied expressions of mortuary behavior are present in other social mammals, and hominins are distinct in the extent of repetition, social collaboration, and shared meaning in the process. Rather than relying on increased encephalization and its relation to complex behavior as a Cognitive Rubicon in human evolution (see Meneganzin and Currie 2022), we suggest that a distinctive cultural, empathetic, collaborative niche dependent on increasingly complex and robust relationships between individuals has also been a primary driver in the development of key aspects of human, or human-like, behavior (Galway-Witham, Cole, and Stringer 2019; Kissel and Fuentes 2021; McBrearty and Brooks 2000; Fuentes 2017; Spikins 2022; DeCasien, Barton, and Higham 2022). The increasing data for complex behavior and meaning-making across the Pleistocene should play a major element in structuring how we investigate, explain, and model the origins and patterns of hominin and human evolution (Kissel and Fuentes 2021; Spikins 2022; Spikins et al. 2019). The current evidence for H. naledi in the Rising Star system pushes back the origins of mortuary, and possibly funerary behaviors, challenges our assumptions about the role and importance of encephalization in human evolution, and suggests that the hominin emotional, socio-cognitive niche is more significant than previously thought.